Characterizing Signal Loss, Error, and Bias in the 21cm Reionization \\Power Spectrum: A Revised Study of PAPER-64

### 2018-03-23 PAPER.tex
Added ACL funding information and tweaked notation in appendix

\section{Acknowledgements}
CC would like to acknowledge the UC Berkeley Chancellor's Fellowship and National Science Foundation Graduate Research 
Fellowship (Division of Graduate Education award 1106400). She would also like to thank Eric Switzer, Miguel Morales, and Bryna Hazelton for helpful discussions. PAPER and HERA 
are supported by grants from the National Science Foundation (awards 1440343, and 1636646). ARP, DCJ, and JEA would 
also like to acknowledge NSF support (awards 1352519, 1401708, and 1455151, respectively). SAK is supported by the University of Pennsylvania School of Arts and Sciences Dissertation Completion Fellowship. JSD acknowledges NSF AAPF
award 1701536. We graciously thank SKA-SA for site infrastructure and observing support.
\label{sec:Ack}

@@

The minimum-variance quadratic estimator $\widehat{p}_\alpha$ for the $\alpha$th bandpower of the power spectrum is given by 
\begin{equation}
\widehat{p}_\alpha = \frac{1} {2 \F_{\alpha \alpha} }\x^t \C^{-1} \Q^{\alpha} \C^{-1} \x,
\end{equation}
where
\begin{equation}
F_{\alpha \alpha} \equiv \frac{1}{2} \textrm{tr} \left( \C^{-1} \Q^\alpha \C^{-1} \Q^\alpha \right)
\end{equation}
is the $\alpha$th diagonal element of the Fisher matrix (for this section only, with no loss of generality, we assume that the data $\textbf{x}$ are real). In our case, however, we do not have \emph{a priori} knowledge of the covariance matrix. Thus, we replace $\C$ with $\Chat$, its data-derived approximation. Our estimator then becomes
\begin{equation}
\label{eq:phatloss}
\widehat{p}_\alpha^\textrm{loss} = \frac{1} {2 \widehat{\F}_{\alpha \alpha} }\x^t \Chat^{-1} \Q^{\alpha} \Chat^{-1} \x,
\end{equation}
where
\begin{equation}
\widehat{F}_{\alpha \alpha} \equiv \frac{1}{2} \textrm{tr} \left( \Chat^{-1} \Q^\alpha \Chat^{-1} \Q^\alpha \right),
\end{equation}
with the label ``loss" to foreshadow the fact that this will be an estimator with signal loss (i.e., a multiplicative bias of less than unity). We will now provide an explicit demonstration of this by modeling the estimated covariance as
\begin{equation}
\label{eq:ChatDef}
\Chat = (1-\eta) \C + \eta \x \x^t,
\end{equation}
where $\eta$ is a parameter quantifying our success at estimating the true covariance matrix. If $\eta = 0$, our covariance estimate has perfectly modeled the true covariance and $\Chat = \C$. On the other hand, if $\eta =1$, then our covariance estimate is based purely on the one realization of the covariance that is our actual data, and we would expect a high level of overfitting and signal loss.

Our strategy for computing the signal loss will be to insert Equation \eqref{eq:ChatDef} into Equation \eqref{eq:phatloss} and to express the resulting estimator $\widehat{p}_\alpha^\textrm{loss}$ in terms of $\widehat{p}_\alpha$. We begin by expressing $\Chat^{-1}$ in terms of $\C^{-1}$ using the Woodbury identity so that
\begin{equation}
\Chat^{-1} = \frac{\C^{-1}}{1-\eta} \left[ \I - \frac{\eta \x \x^t \C^{-1}}{1+ \eta (g-1)}\right],
\end{equation}
where we have defined $g \equiv \x^t \C^{-1} \x$. Inserting this into our Fisher estimate we have
\begin{equation}
\widehat{F}_{\alpha \alpha} = \frac{F_{\alpha \alpha}}{(1-\eta)^2} \left[ 1 -\frac{\eta }{1+ \eta (g-1)} \frac{h_{\alpha \alpha}}{F_{\alpha \alpha}} + \frac{1}{2} \left( \frac{\eta }{1+ \eta (g-1)} \right)^2 \frac{h_\alpha^2}{F_{\alpha \alpha}}\right],
\end{equation}
where $h_\alpha \equiv \x^t \C^{-1} \Q^\alpha \C^{-1} \x $ and $h_{\alpha \alpha} \equiv \x^t \C^{-1} \Q^\alpha \C^{-1} \Q^\alpha \C^{-1}\x $. Note that $g$, $h_\alpha$, and $h_{\alpha \alpha}$ are all random variables, since they depend on $\x$. Inserting these expressions into our estimator gives
\begin{equation}
\label{eq:phatlossexpanded}
\widehat{p}_\alpha^\textrm{loss} = \frac{1}{2} \frac{h_\alpha}{F_{\alpha \alpha}} \left[ 1 - \frac{\eta g}{1+ \eta (g-1)}\right]^2  \left[ 1 -\frac{\eta }{1+ \eta (g-1)} \frac{h_{\alpha \alpha}}{F_{\alpha \alpha}} + \frac{1}{2} \left( \frac{\eta }{1+ \eta (g-1)} \right)^2 \frac{h_\alpha^2}{F_{\alpha \alpha}}\right]^{-1}.
\end{equation}
Both for the purposes of analytical tractability and to provide intuition, we expand this expression to leading 
order in $\eta$. This approximates the limiting case where the covariance $\Chat$ is close to the ideal and the 
lossy covariance is a small perturbation.  The result is
\begin{equation}
\widehat{p}_\alpha^\textrm{loss} \approx \frac{1}{2} \frac{h_\alpha}{F_{\alpha \alpha}} \left[ 1 - \eta \left( g - \frac{h_{\alpha \alpha}}{F_{\alpha \alpha}}\right)\right].
\end{equation}
Taking the ensemble average of both sides and noting that the true power spectrum $p_\alpha$ is equal to $\langle h_\alpha \rangle / 2 F_{\alpha \alpha}$, we obtain
\begin{equation}
\langle \widehat{p}_\alpha^\textrm{loss} \rangle \approx (1- \eta N) p_\alpha + 4 \eta \frac{\rm{tr} (\C^{-1} \Q^\alpha \C^{-1} \Q^\alpha \C^{-1} \Q^\alpha )}{\left[ \rm{tr} (\C^{-1} \Q^\alpha \C^{-1} \Q^\alpha  ) \right]^2} \approx (1- \eta N) p_\alpha,
\end{equation}
where $N$ is the length of $\x$. In the last step we dropped the final term, since it scales as $\eta p_\alpha$ (without the factor of $N$) and is therefore typically small compared to the terms that have been retained.

>>>>

\section{Acknowledgements}
CC would like to acknowledge the UC Berkeley Chancellor's Fellowship and National Science Foundation Graduate Research 
Fellowship (Division of Graduate Education award 1106400). She would also like to thank Eric Switzer, Miguel Morales, and Bryna Hazelton for helpful discussions. PAPER and HERA 
are supported by grants from the National Science Foundation (awards 1440343, and 1636646). ARP, DCJ, and JEA would 
also like to acknowledge NSF support (awards 1352519, 1401708, and 1455151, respectively). SAK is supported by the University of Pennsylvania School of Arts and Sciences Dissertation Completion Fellowship. JSD acknowledges NSF AAPF
award 1701536. AL acknowledges support for this work by NASA through Hubble Fellowship grant \#HST-HF2-51363.001-A awarded by the Space Telescope Science Institute, which is operated by the Association of Universities for Research in Astronomy, Inc., for NASA, under contract NAS5-26555. We graciously thank SKA-SA for site infrastructure and observing support.
\label{sec:Ack}

@@

The minimum-variance quadratic estimator $\widehat{p}_\alpha$ for the $\alpha$th bandpower of the power spectrum is given by 
\begin{equation}
\widehat{p}_\alpha = \frac{1} {2 \F_{\alpha \alpha} }\x^t \C^{-1} \Q^{\alpha} \C^{-1} \x,
\end{equation}
where
\begin{equation}
F_{\alpha \alpha} \equiv \frac{1}{2} \textrm{tr} \left( \C^{-1} \Q^\alpha \C^{-1} \Q^\alpha \right)
\end{equation}
is the $\alpha$th diagonal element of the Fisher matrix. For this section only, with no loss of generality, we assume that the data $\textbf{x}$ are real. We also assume for simplicity that $\mathbf{x}$ is the data from a single instant in time, so that it is of length $N_f$, where $N_f$ is the number of frequency channels.

In our case, we do not have \emph{a priori} knowledge of the covariance matrix. Thus, we deviate from the true minimum-variance quadratic estimator and replace $\C$ with $\Chat$, its data-derived approximation. Our estimator then becomes
\begin{equation}
\label{eq:phatloss}
\widehat{p}_\alpha^\textrm{loss} = \frac{1} {2 \widehat{\F}_{\alpha \alpha} }\x^t \Chat^{-1} \Q^{\alpha} \Chat^{-1} \x,
\end{equation}
where
\begin{equation}
\widehat{F}_{\alpha \alpha} \equiv \frac{1}{2} \textrm{tr} \left( \Chat^{-1} \Q^\alpha \Chat^{-1} \Q^\alpha \right),
\end{equation}
with the label ``loss" to foreshadow the fact that this will be an estimator with signal loss (i.e., a multiplicative bias of less than unity). We will now provide an explicit demonstration of this by modeling the estimated covariance as
\begin{equation}
\label{eq:ChatDef}
\Chat = (1-\eta) \C + \eta \x \x^t,
\end{equation}
where $\eta$ is a parameter quantifying our success at estimating the true covariance matrix. If $\eta = 0$, our covariance estimate has perfectly modeled the true covariance and $\Chat = \C$. On the other hand, if $\eta =1$, then our covariance estimate is based purely on the one realization of the covariance that is our actual data, and we would expect a high level of overfitting and signal loss.

Our strategy for computing the signal loss will be to insert Equation \eqref{eq:ChatDef} into Equation \eqref{eq:phatloss} and to express the resulting estimator $\widehat{p}_\alpha^\textrm{loss}$ in terms of $\widehat{p}_\alpha$. We begin by expressing $\Chat^{-1}$ in terms of $\C^{-1}$ using the Woodbury identity so that
\begin{equation}
\Chat^{-1} = \frac{\C^{-1}}{1-\eta} \left[ \I - \frac{\eta \x \x^t \C^{-1}}{1+ \eta (g-1)}\right],
\end{equation}
where we have defined $g \equiv \x^t \C^{-1} \x$. Inserting this into our Fisher estimate we have
\begin{equation}
\widehat{F}_{\alpha \alpha} = \frac{F_{\alpha \alpha}}{(1-\eta)^2} \left[ 1 -\frac{\eta }{1+ \eta (g-1)} \frac{h_{\alpha \alpha}}{F_{\alpha \alpha}} + \frac{1}{2} \left( \frac{\eta }{1+ \eta (g-1)} \right)^2 \frac{h_\alpha^2}{F_{\alpha \alpha}}\right],
\end{equation}
where $h_\alpha \equiv \x^t \C^{-1} \Q^\alpha \C^{-1} \x $ and $h_{\alpha \alpha} \equiv \x^t \C^{-1} \Q^\alpha \C^{-1} \Q^\alpha \C^{-1}\x $. Note that $g$, $h_\alpha$, and $h_{\alpha \alpha}$ are all random variables, since they depend on $\x$. Inserting these expressions into our estimator gives
\begin{equation}
\label{eq:phatlossexpanded}
\widehat{p}_\alpha^\textrm{loss} = \frac{1}{2} \frac{h_\alpha}{F_{\alpha \alpha}} \left[ 1 - \frac{\eta g}{1+ \eta (g-1)}\right]^2  \left[ 1 -\frac{\eta }{1+ \eta (g-1)} \frac{h_{\alpha \alpha}}{F_{\alpha \alpha}} + \frac{1}{2} \left( \frac{\eta }{1+ \eta (g-1)} \right)^2 \frac{h_\alpha^2}{F_{\alpha \alpha}}\right]^{-1}.
\end{equation}
Both for the purposes of analytical tractability and to provide intuition, we expand this expression to leading 
order in $\eta$. This approximates the limiting case where the covariance $\Chat$ is close to the ideal and the 
lossy covariance is a small perturbation.  The result is
\begin{equation}
\widehat{p}_\alpha^\textrm{loss} \approx \frac{1}{2} \frac{h_\alpha}{F_{\alpha \alpha}} \left[ 1 - \eta \left( g - \frac{h_{\alpha \alpha}}{F_{\alpha \alpha}}\right)\right].
\end{equation}
Taking the ensemble average of both sides and noting that the true power spectrum $p_\alpha$ is equal to $\langle h_\alpha \rangle / 2 F_{\alpha \alpha}$, we obtain
\begin{equation}
\langle \widehat{p}_\alpha^\textrm{loss} \rangle \approx (1- \eta N_f) p_\alpha + 4 \eta \frac{\rm{tr} (\C^{-1} \Q^\alpha \C^{-1} \Q^\alpha \C^{-1} \Q^\alpha )}{\left[ \rm{tr} (\C^{-1} \Q^\alpha \C^{-1} \Q^\alpha  ) \right]^2} \approx (1- \eta N_f) p_\alpha,
\end{equation}
where recall that $N_f$ is the length of $\x$, or the number of frequency channels. In the last step we dropped the final term, since it scales as $\eta p_\alpha$ (without the factor of $N$) and is therefore typically small compared to the terms that have been retained.
